Type Visualization for R

### 2020-06-20 paper/main.tex
paper: Cites

\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling useful
  information from programs remain difficult. Important dimensions of
  code, like run-time type data, are inscrutable without appropriate tools.
  We built {\sc TypeVis} to explore and analyze run-time types in
  the R programming language by means of visualization.
  Our system focuses on understanding and comparing the type signatures of
  function calls across the R ecosystem.
  {\sc TypeVis} uses flows to intuitively represent categorical
  and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing
  language design decisions---specifically of a gradual type system
  currently being developed for R.

  \todo{Fill in keywords below.}
}

@@

\CCScatlist{
%  \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visu\-al\-iza\-tion techniques}{Treemaps};%  \CCScatTwelve{Human-centered computing}{Visu\-al\-iza\-tion}{Visualization design and evaluation methods}{}
}

%\CCScatlist{
  %\CCScat{H.5.2}{User Interfaces}{User Interfaces}{Graphical user interfaces (GUI)}{};
  %\CCScat{H.5.m}{Information Interfaces and Presentation}{Miscellaneous}{}{}
%}

@@

%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.

%% the only exception to this rule is the \firstsection command
\firstsection{Introduction}

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead

@@

We are not aware of any previous work that applies
visualization techniques for the purposes of programming language design.
However, there are existing publications on understanding
software systems through visualization.
Bohnet visualize massive program execution traces
to assist developers in code comprehension of large software systems.
Their tool applies pruning and summarization to reduce trace size.
The visualization focuses on call topology,
describing the structure of function invocations,
rather than types.
Telea also develop a methodology for displaying call structure.
However, they visualize call graphs of systems in their entirety,
rather than individual execution traces.
Our visualization can be thought of as a hybrid of the two,
investigating the system as a whole,
but doing so via execution traces.
In addition to standard call graphs,
LaToza adorn nodes and edges with extra information,
allowing the user gain further understanding of the system
under examination.
This includes indicating methods that come from the same class---a
basic form of type information.

@@

These goals can be classified using an abstract task analysis framework.
See table~\ref{tab:tasks} for a categorization of these domain tasks
into Munzner's mid-level and low-level abstract task taxonomy. \todo{Cite Munzner}

@@

Treemaps, introduced by Shneiderman,
are often used for displaying hierarchical data.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
R packages are not structured deeply
so the treemaps in our data are only one
level deep.
However, if the dataset was more deeply nested,
{\sc TypeVis} can be configured to
handle that.

@@

Prominently featured as the center visualization,
a parallel sets diagram encodes type signatures of the selected
functions as flows.
Parallel sets,
proposed by Kosara,
are ideally suited for type signatures.
Each data point is can be thought of as
a tuple $(f, \tau_1, \ldots, \tau_n)$
corresponding to the function name,
argument types,
and return type.
Every component is categorical,
and we are interested in visualizing frequencies of these points.
This is the exact use-case for parallel sets.

@@

\todo{Finish these.}
\acknowledgments{
  The authors wish to thank the following people:
  students of CS 7250 for useful design feedback,
  contributors to all the open source projects that {\sc TypeVis} depends on.
  This work was supported in part by a Northeastern Graduate Fellowship.
}

>>>>

\abstract{
  Data-driven approaches to programming language design are uncommon.
  Despite the availability of large code repositories, distilling useful
  information from programs remain difficult. Important dimensions of
  code, like run-time type data, are inscrutable without appropriate tools.
  We built {\sc TypeVis} to explore and analyze run-time types in
  the R programming language by means of visualization.
  Our system focuses on understanding and comparing the type signatures of
  function calls across the R ecosystem.
  {\sc TypeVis} uses flows to intuitively represent categorical
  and numerical aspects of type signatures.
  Insights derived from our visualization are aimed at informing
  language design decisions---specifically of a gradual type system
  currently being developed for R.
}

@@

\CCScatlist{
  \CCScatTwelve{Human-centered computing}{Visualization}{Visualization systems and tools}{}
}

@@

\firstsection{Introduction}

@@

@@

We are not aware of any previous work that applies
visualization techniques for the purposes of programming language design.
However, there are existing publications on understanding
software systems through visualization.
Bohnet et al.~\cite{bohnet:2009} visualize massive program execution traces
to assist developers in code comprehension of large software systems.
Their tool applies pruning and summarization to reduce trace size.
The visualization focuses on call topology,
describing the structure of function invocations,
rather than types.
Telea et al.~\cite{telea:2009} also develop a methodology for displaying call structure.
However, they visualize call graphs of systems in their entirety,
rather than individual execution traces.
Our visualization can be thought of as a hybrid of the two,
investigating the system as a whole,
but doing so via execution traces.
In addition to standard call graphs,
LaToza and Myers~\cite{latoza:2011} adorn nodes and edges with extra information,
allowing the user gain further understanding of the system
under examination.
This includes indicating methods that come from the same class---a
basic form of type information.

@@

These goals can be classified using an abstract task analysis framework.
See table~\ref{tab:tasks} for a categorization of these domain tasks
the mid-level and low-level abstract task taxonomy
established by Brehmer and Munzner~\cite{brehmer:2013}.

@@

Treemaps are often used for displaying hierarchical data~\cite{shneiderman:1992}.
Package systems in many languages,
like Java,
contain deeply nested namespace hierarchies.
This idiom is appropriate for navigation of such
hierarchies.
R packages are not structured deeply
so the treemaps in our data are only one
level deep.
However, if the dataset was more deeply nested,
{\sc TypeVis} can be configured to
handle that.

@@

Prominently featured as the center visualization,
a parallel sets diagram encodes type signatures of the selected
functions as flows.
Parallel sets,
proposed by Kosara et al.~\cite{kosara:2006},
are ideally suited for type signatures.
Each data point is can be thought of as
a tuple $(f, \tau_1, \ldots, \tau_n)$
corresponding to the function name,
argument types,
and return type.
Every component is categorical,
and we are interested in visualizing frequencies of these points.
This is the exact use-case for parallel sets.

@@

\acknowledgments{
  The authors wish to thank the students of CS 7250 for useful design feedback,
  and contributors to all the open source projects that {\sc TypeVis} depends on.
  This work was supported in part by a Northeastern Graduate Fellowship.
}
